Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-04 ind_files/03_BD_processes.tex
Change filename of chap4 to examples.tex

where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration. For most of our biological cases, both of these conditions will be satisfied, and so we assume going further that all our SDEs are always well defined and admit solutions.} and $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that is only seen in finite populations. Mathematically, this term acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a decrease in frequency - notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why the former routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\

>>>>

where $F(t,x)$ and $G(t,x)$ are `nice' functions\footnote{For the mathematically oriented reader, there are two requirements: Firstly, we require the functions to have `linear growth', meaning that we can find a constant $C > 0$ such that $\|F(t,x)\| + \|G(t,x)\| \leq C(1+\|x\|)$ for every $x \in \mathbb{R}^{d}$ and $t > 0$. We also require `Lipschitz continuity', which means that we can find a constant $L > 0$ such that $\|F(t,x) - F(t,y)\| + \|G(t,x)-G(t,y)\| \leq L\|x-y\|$ for every pair $x,y \in \mathbb{R}^d$ and $t > 0$. Here, $\|\cdot\|$ denotes the natural norm on the space under consideration and for our cases will usually be the Euclidean norm. For biological systems, both of these conditions will usually be satisfied, and so we assume going further that all our SDEs are always well-defined and admit solutions.} In the physics literature, $F$ and $G$ are often called the `drift' and `diffusion' of the process respectively. However, we will not use this terminology here due to  potential confusion with genetic drift (which actually corresponds to the `diffusion' in the physics terminology). $B_t$ is the so-called `standard Brownian motion'. Named after the botanist Robert Brown (who was looking at the random erratic motion of pollen grains in water under a microscope), $\{B_t\}_{t \geq 0}$ is a stochastic process that is supposed to model `random noise' or `undirected diffusion' of a particle. If one imagines $B_t$ as recording the position of a small pollen grain at time $t$, then $B_t$ can be formally thought of as a process that has the following properties:

@@

where $B^{(1)}_t,B^{(2)}_t, \ldots, B^{(m)}_t$ are $m$ independent one-dimensional standard Brownian motion processes. The first term in this expression is the same as equation \eqref{nD_replicator_mutator} and describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation. We saw in the previous section that the replicator-mutator equation is generically satisfied in the infinite population limit for all birth-death processes that are in the (very general) functional forms \eqref{nD_functional_forms_for_replicator}, a result that is made explicit in \eqref{nD_eqn_for_frequencies} (just take $K \to \infty$ to convince yourself). The second term is an additional directional force on population composition that is only seen in finite populations. Mathematically, this term acts exactly analogously to the replicator equation but in the opposite direction (A higher relative $\tau_i$ leads to a decrease in frequency - notice the minus sign before the second term in \eqref{nD_eqn_for_frequencies}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase (such as cheaters and highly competitive species), which could explain why the former routinely persist for long periods in finite population models despite extinction being predicted in infinite population models. The fact that the entire term is multiplied by $(KN_K(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\

### 2023-02-04 ind_files/04_examples.tex
Change filename of chap4 to examples.tex

The methods outlined in the above section have recently been used to study the population dynamics of a finite population playing a so-called `matrix game' (An evolutionary game for which you can write down a payoff matrix) with 2 pure strategies \citep{tao_stochastic_2007}. Based on the interpretation of what each type represents, this is mathematically equivalent to studying frequency-dependent selection on a one-locus two-allele gene (with a bijective genotype-phenotype map and no mutations) or studying two-species competitive Lotka-Volterra dynamics, as we will show below.\\

@@

which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. We briefly note here that if $m$ is large, then both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation. If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, then, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus reducing to an $m-1$ dimensional system of equations and obtaining the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics. However, we will not explore this further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} for a review of the ideas of dynamics on slow manifolds.\\

>>>>

The methods outlined in the above section have recently been used to study the population dynamics of a finite population playing a so-called `matrix game' (An evolutionary game for which you can write down a payoff matrix) with 2 pure strategies \citep{tao_stochastic_2007}. Based on the interpretation of what each type represents, this is mathematically equivalent to studying frequency-dependent selection on a one-locus two-allele gene (with a bijective genotype-phenotype map and no mutations) or studying two-species competitive Lotka-Volterra dynamics, as we will show below. The stochastic Lotka-Volterra competition model shown below has also been proved to be equivalent to an $m$-allele Moran model under certain limits \citep{constable_mapping_2017}.\\

@@

which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $B$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. We briefly note here that if $m$ is large, then both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation. If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, then, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics \citep{constable_mapping_2017,parsons_dimension_2017}. However, we will not explore this further in this manuscript, and refer the reader to \citep{constable_stochastic_2013} and \citep{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.\\
